---
title: "Interesting R Examples"
author: Ilker Kurtulus - IE582 - Fall 2018
---
```{r,echo=FALSE,results="hide"}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

```{r,echo=FALSE,eval=FALSE}
rmarkdown::render("/home/baydogan/Courses/IE582/Fall18/Guidelines/example_homework_0.Rmd",output_format="pdf_document")
rmarkdown::render("/home/baydogan/Courses/IE582/Fall18/Guidelines/example_homework_0.Rmd",output_format="html_document")
```

1. [Exploring Survival on the Titanic](https://www.kaggle.com/mrisdal/exploring-survival-on-the-titanic)

    This example is important because it is an R Notebook on Titanic data set. Titanic is considered as one of the Data Science 101 data set and this notebook provides a good exploratory data analysis. It is also a Kaggle Kernel, which is a very good source for this kind of example R codes. The notebook also covers feature engineering, missing data imputation and modeling.

2.Read Data
```{r eval=TRUE}
df = read.csv('/Users/ilkerkurtulus/Documents/cse-master/ie582/data/train.csv')
library(stringr)
x = str_split_fixed(df$Name, ", ", 2)
y = str_split_fixed(x[,2], ". ", 2)
df["title"] = y[,1]
head(df)
```
