---
title: "Interesting R Examples"
author: Ilker Kurtulus - IE582 - Fall 2018
---
```{r,echo=FALSE,results="hide"}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

```{r,echo=FALSE,eval=FALSE}
rmarkdown::render("/home/baydogan/Courses/IE582/Fall18/Guidelines/example_homework_0.Rmd",output_format="pdf_document")
rmarkdown::render("/home/baydogan/Courses/IE582/Fall18/Guidelines/example_homework_0.Rmd",output_format="html_document")
```


1. While discovering titanic dataset i encountered that reading data stored in local to the github is a bit tricky.
To do that one option is the following:
+ Save your data in local git directory then commit and push so that your file will be on your gitrepo.
+ Read your file from raw.githubusercontent link which can be obtained by clicking raw button in above data storage directory.
```{r eval=TRUE}
library(stringr)
df = read.csv("https://raw.githubusercontent.com/BU-IE-582/fall18-ilkerkurtulus/master/files/train.csv")
head(df)
```

2. A basic string operation for feature engineering by using library called stringr.

```{r eval=TRUE}
library(stringr)
x = str_split_fixed(df$Name, ", ", 2)
y = str_split_fixed(x[,2], ". ", 2)
df["title"] = y[,1]
head(df)
```

3. More solutions can be found in [here]("https://blogdown-demo.rbind.io/2018/02/27/r-file-paths/") where you can find additional info about r file paths.

4. Another example is a kaggle post which describes constructing Neural Networks within different type of predictors such that text, image, numerical and categorical. Its a good example for data preprocessing.  ["Dance with Ensemble" Sharing Thread]("https://www.kaggle.com/c/avito-demand-prediction/discussion/59880")

5. This is one of my favorite data visualization post in Kaggle. It contains multiple visualization libraries.  [Be my guest - Recruit Restaurant EDA]("https://www.kaggle.com/headsortails/be-my-guest-recruit-restaurant-eda").

